Fractional Types

### 2019-11-22 fracGC/p.tex
trace 2 & 3

The paper includes a number of languages that express different
points. The three most important languages and their relationships are
highlighted in this diagram:
\begin{center}
\input{lang.tikz}
\end{center}

@@

%%%%%
\subsection{Choosing Ancillae Values: $\Pi/D$}

@@

We will call this language $\Pi/D$ for the fractional extension of
$\Pi$ with a dynamic check. In addition to being more expressive than
the language in the previous subsection, $\Pi/D$ has the advantage of
replacing the cumbersome cardinality-based evidence that the type is
not empty by the more intuitive evidence of choosing a value of the
type. The simplications are evident in the Agda formalization:

@@

\begin{wrapfigure}{r}{0.2\textwidth}
  \input{parity.tikz}
\end{wrapfigure}
We illustrate the new style of ancilla management with an interesting
example used in quantum circuits as a dynamic assertion of
entanglement~\cite{DBLP:journals/cal/ZhouB19}:

\PIFDparity{}

@@

%%
\subsection{Abstract Machine for $\Pi$}

@@

%%
\subsection{Abstract Machine for $\Pi/D$}

@@

\begin{figure}[htb]
\includegraphics[scale=0.4]{trace2.png}
\caption{Execution trace of \AgdaFunction{id'}}
\label{fig:trace-id}
\end{figure}

@@

%%%%%
\subsection{Pointed and Singleton Types: $\Pi/\bullet$}

Following the development for the toy language above, we create a
lifted version of fractional $\Pi$ with pointed types which we call
$\Pi/\bullet$. In addition to the explicit pointed types of the form
$t$\Acon{\#}$v$, we will need a multiplicative structure over pointed
types, fractional types, and a special kind of pointed type that
includes just one value, a singleton type. The singleton types will
allow the type system to track the flow of one particular value, which
is exactly what is needed to prove the safety of deallocation:

@@

The combinators in the lifted language $\Pi/\bullet$ consist of all the combinators
in the core $\Pi$ language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract} which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

\includegraphics[scale=0.5]{trace3.png}

@@

By lifting programs and their evaluation to the type level, we can
naturally leverage the typechecking process to verify properties of
interest, including the safe de-allocation of ancillae. One ``could''
just write programs in that style in $\Pi/\bullet$: indeed the dual nature of proofs
and programs is more and more exploited in languages like the one used
to formalize this paper (Agda). The examples in Sec.~\ref{sec:cat} are
written in that style and they are relatively similar to programs
in the ``normal'' $\Pi/D$ language.

However, it is also often the case than constructive proofs are
further processed to extract a native efficient programs that eschew
the overhead of maintaining information needed just for proof
invariants. In our case, the question is whether we can extract from
$\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
system, a simpler runtime representation, and yet is guaranteed to be
safe and hence can run without the runtime checks associated with
de-allocation sites. In this section, we show that this indeed the
case.

% First we can inject plain Pi type and value into PiFracDyn:
% %% \ednote{Pi/.U is not defined before its use below}
% \INJU{}
% and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
% \INJcomb{}
% The given injection is correct because it does not change the denotation of type
% \INJUeq{}
% ,the value
% \INJVeq{}
% and the interpretation of PiFracDyn is consistent with Pi:
% \INJEvaleq{}

The extraction map is fully implemented in the underlying Agda
formalization. We present the most significant highlights. There are
three important functions who signatures are below:

\EXTsig{}

The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
and a value in the type. The function \Afun{Ext∙⟶} maps a
$\Pi/\bullet$ combinator to a $\Pi/D$ combinator. And finally the
function \Afun{Ext≡} asserts that extracted code cannot throw an
exception (it must return \Acon{just}\ldots).

Each of these functions has one or two enlightening cases which we
explain:

In $\Pi/D$ the fractional type expresses that it expects a particular
value but lacks any mechanisms to enforce this requirement. Thus we
have no choice when mapping a fractional type from $\Pi/\bullet$ to 
$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} type:

\EXTu{}

% There is no corresponding type of Recip
% in plain Pi, but we can map it to the
% \AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
% type in PiFracDyn.  Note that for all other type the extracted value
% is erased from type except for
% \AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
% which we still keep it in type.

When mapping $\Pi/\bullet$ combinators to 
$\Pi/D$ combinators, the main interesting cases are for \Acon{η} and
\Acon{ε}. In those we just use the values from the pointed type as
choices for the ancilla value and expectation for the gc process:

\EXTcomb{}

% Once we extract the
% corresponding type we can extract the combinator, the combinators
% about monad/comonad just becomes \AgdaInductiveConstructor{id↔} since
% the information in type is erased, for the rest combinators there
% exist direct corresponding combinators in PiFracDyn:

% Finally we can prove the correctness of the extraction, the lifted
% multiplicative combinators have the corresponding combinators in
% PiFracDyn which represent the same computation so the proof is just
% \AgdaInductiveConstructor{refl}.  Because the combinators about
% monad/comonad does not really computes so for those case the proof is
% also \AgdaInductiveConstructor{refl}.  And for
% \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.  The
% interesting case is \AgdaInductiveConstructor{ε} where the singleton
% type guarantee that the dynamic check won't fail.

Finally we can prove the correctness of the extraction. The punchline
is in the following case:

\EXTeq{}

In this case, the singleton type in $\Pi/\bullet$ guarantees that the
runtime check cannot fail.

@@

In addition to the examples presented throughout the paper, we present
examples that provides novel perspectives on gc processes, first as
space transformers, and second as tracing operators. Both ideas are
inspired from known constructions in compact closed
categories~\cite{dblp:conf/calco/abramsky05}.

@@

As the diagram below (take from Abramsky's
paper~\cite{dblp:conf/calco/abramsky05}) illustrates, this is not a
trivial construction in the sense that the permutation from $A$ to $B$
uses $C$ in non-trivial ways.

@@

There has been ample work on ancillae before us.  For example
\cite{PhysRevA.94.042337} explicitly represent the lifetime of
quantum wires, enable quantum wires to be recycled, leading
to more efficient circuits.  Interesting applications of
ancillae is in the recent paper~\cite{DBLP:journals/cal/ZhouB19}
on Quantum runtime assertions.  In section~\ref{sub:id}, we
already mentioned dynamic assertion of entanglement; they
also show how to provide (dynamic) assertions on classical
states, and testing of superposition states.  Our techniques
should combine with theirs to provide even more possibilities
for dynamic assertions. Ancilla management is a complex
business, and the verified circuit compiler
ReVerC~\cite{amy2017verified} goes to great pains to
ensure that ancilla are dealt with properly.  Both
QCL~\cite{omer2002procedural} and Quipper~\cite{Green:2013:QSQ:2491956.2462177}
also do automatic (but not certified) ancilla management during code
generation.  This remains an active area of research, as
outlined in the CACM paper~\cite{valiron2015programming}.

@@

Another different thread of work stems from categorical models
of quantum computation, especially those based on dagger
compact closed categories as introduced
in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
expanded upon in~\cite{abramsky2009categorical}.  Various
results can be obtained in that general setting, such as the
no-cloning theorem~\cite{abramsky2009no}. The $2016$ book
\cite{chiribella2016quantum} is a good survey of the
developments since. Selinger~\cite{selinger2007dagger,Selinger:2011:FDH:1942319.1942398}
has been particularly active in this area. Of particular
note for our applications are two 2005 papers by Abramsky
\cite{abramsky2005structural,dblp:conf/calco/abramsky05}
with explicit examples of circuits using trace that are very useful tests
for our approach.

Also related are various approaches to quantum lambda
calculi, such as~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}.

>>>>

@@

%%%%%
\subsection{Choosing Ancillae Values}

@@

In addition to being more expressive, this version has the advantage
of replacing the cumbersome cardinality-based evidence that the type
is not empty by the more intuitive evidence of choosing a value of the
type. The simplications are evident in the Agda formalization:

@@

We illustrate the new style of ancilla management with an interesting
example used in quantum circuits as a dynamic assertion of
entanglement~\cite{DBLP:journals/cal/ZhouB19}:

\begin{wrapfigure}[0]{r}{0.2\textwidth}
  \input{parity.tikz}
\end{wrapfigure}
\PIFDparity{}

@@

%%
\subsection{Abstract Machine without Fractionals}

@@

%%
\subsection{Abstract Machine with Fractionals}

@@

{\center\input{trace2.tex}}

@@

%%%%%
\subsection{Pointed and Singleton Types}

Following the development for the toy language above, we create a
lifted version of fractional $\Pi$ with pointed types. In addition to
the explicit pointed types of the form $t$\Acon{\#}$v$, we will need a
multiplicative structure over pointed types, fractional types, and a
special kind of pointed type that includes just one value, a singleton
type. The singleton types will allow the type system to track the flow
of one particular value, which is exactly what is needed to prove the
safety of deallocation:

@@

The combinators in the lifted language consist of all the combinators
in the base language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract} which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

{\center\input{trace3.tex}}

@@

Exploit dependent types to reify proofs in the type system. Type
checking requires (partial) eval. Once we have a proof can we somehow
go back to \verb|PiFracDyn| and remove the runtime check?

Polymorphic type for ancilla?
\\
First we can inject plain Pi type and value into PiFracDyn:
%% \ednote{Pi/.U is not defined before its use below}
\INJU{}
and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
\INJcomb{}
The given injection is correct because it does not change the denotation of type
\INJUeq{}
,the value
\INJVeq{}
and the interpretation of PiFracDyn is consistent with Pi:
\INJEvaleq{}
To extract a combinator from PiPointedFrac, first we need to extract the corresponding type for PiFracDyn.
Since every type (except Recip) in PiPointedFrac is pointed type so we can always extract a type for plain
Pi and a value from it.
There is no corresponding type of Recip in plain Pi, but we can map it to the
\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v} type in PiFracDyn.
Note that for all other type the extracted value is erased from type except for
\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
which we still keep it in type.
\EXTU{}
Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased,
for the rest combinators there exist direct corresponding combinators in PiFracDyn:
\EXTUComb{}
Finally we can prove the correctness of the extraction,
the lifted multiplicative combinators have the corresponding combinators in PiFracDyn which represent
the same computation so the proof is just \AgdaInductiveConstructor{refl}.
Because the combinators about monad/comonad does not really
computes so for those case the proof is also \AgdaInductiveConstructor{refl}.
And for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
the dynamic check won't fail.
\EXTeq{}

@@

In addition to the examples presented throughout the paper, we present
examples that provides novel perspectives on gc processes, first as
space transformers, and second as tracing operators. Both ideas are
inspired from known constructions in compact closed
categories~\cite{DBLP:conf/calco/Abramsky05}.

@@

As the diagram below (take from Abramsky's
paper~\cite{DBLP:conf/calco/Abramsky05}) illustrates, this is not a
trivial construction in the sense that the permutation from $A$ to $B$
uses $C$ in non-trivial ways.

@@

There has been ample work on ancillae before us.  For example%
\cite{PhysRevA.94.042337} explicitly represent the lifetime of
quantum wires, enable quantum wires to be recycled, leading
to more efficient circuits.  Interesting applications of
ancillae is in the recent paper~\cite{DBLP:journals/cal/ZhouB19}
on Quantum runtime assertions.  In section~\ref{sub:id}, we
already mentionned dynamic assertion of entanglement; they
also show how to provide (dynamic) assertions on classical
states, and testing of superposition states.  Our techniques
should combine with theirs to provide even more possibilities
for dynamic asserctions. Ancilla management is a complex
business, and the verified circuit compiler
ReVerC~\cite{amy2017verified} goes to great pains to
ensure that ancilla are dealt with properly.  Both
QCL~\cite{omer2002procedural} and Quipper~\cite{Green:2013:QSQ:2491956.2462177}
also do automatic (but not certified) ancilla management during code
generation.  This remains an active area of research, as
outlined in the CACM paper~\cite{valiron2015programming}.

@@
